### 2024-04-14 docs/SOSP 2024/caching.tex
Fix compile errors. Add latest pdf
Edit caching section

\section{Representations}
\label{sec:representations}

The design of the \Treebeard{} compiler allows the implementation of different strategies for 
the in-memory representation of the model. The compiler currently has implementations for 
the three representations shown in Figure \ref{Fig:Representations}. The array and sparse 
representations are the ones described in the \Treebeard{} paper\cite{Treebeard}. The reorg 
representation is the representation used by the RAPIDs library\cite{RAPIDs}. 
The \textbf{array representation} is the simplest representation where the trees are stored in an array
in level order. The \textbf{sparse representation} stores the trees in a sparse format where only the
non-zero nodes are stored and nodes contain pointers to their children. The \textbf{reorg representation}
interleaves the array representation of each tree in the model: all root nodes are stored first, then 
the left children of all the roots and so on.
\begin{figure}[htb]
  \centering
  \includegraphics[width=\linewidth]{figures/representations.png}
  \caption{The three representations supported by \Treebeard{}.}
  \label{Fig:Representations}
\end{figure}

One of the major changes we make to the original design of \Treebeard{} \cite{Treebeard} is to
separate the implementation of representations from the rest of the compiler. This allows
us to implement representations as plugins to the compiler. We do this by defining an interface 
that each representation needs to implement. The code generator is implemented using methods 
on the representation interface thus hiding details of the actual representation from the core 
compiler. The interface abstracts the following details. 
\begin{itemize}
  \item \textbf{Buffer allocation:} The representation object exposes methods that generate 
  buffer allocations in the IR. 
  \item \textbf{Moving to child nodes:} Given the value of the predicate at a node (or tile),
  generate code to move to the appropriate child node of the current node.
  \item \textbf{Details of leaves:} The interface abstracts determining whether the current
  node is a leaf (which is needed to terminate walks) and how to get the value of the leaf
  (which is needed to get a tree's prediction).
  \item \textbf{Caching trees:} Since reading the trees into shared memory on GPU (or prefetching)
  them on CPU require knowledge of the buffer layout, the task of generating caching code for trees
  is delegated to the representation object. 
  \item \textbf{Loading thresholds and feature indices:} The representation object provides MLIR 
  rewrite patterns to lower operations that load thresholds and feature indices to LLVM IR. This
  is necessary because the details of how to load the thresholds and feature indices from the model 
  buffers is representation-specific.
\end{itemize}

In summary, the representation interface abstracts the details of how the model is stored in memory
and allows the compiler to generate code without having to explicitly know the details of the
representation. This design allows us to implement new representations without changing the core
compiler infrastructure. Implementing the representations as plugins also allows us to reuse
the implementations across different lowering pipelines. For example, the code for the array 
and sparse representations is almost fully shared between the CPU and GPU lowering pipelines.

\section{Caching Implementation}
\label{sec:caching}

@@

\Treebeard{} provides mechanisms to cache both trees and input rows 
on both the CPU and GPU. As described in Section \ref{sec:schedule}, 
the user can specify that the working set of an iteration of an index
variable needs to be cached using the \op{Cache()} directive. This 
provides a unified way to specify caching of both trees and input rows.

\Treebeard{} implements caching at the granularity of a tree or a row.
Also, the semantics of caching depends on the target processor. For the CPU,
caching is implemented as prefetching, while for the GPU, caching is implemented
using shared memory.

\subsection{IR Representation of Caching}
Caching is encoded in the mid-level IR using the \op{cacheTrees} and \op{cacheRows} operations. 
These operations are generated when the HIR lowered to MIR and \op{Cache()} is specified on an 
index variable in the schedule. While the HIR is being lowered and a cached index variable is 
encountered, the compiler generates a \op{cacheTrees} or \op{cacheRows} operation depending on 
whether the index variable is a tree or a batch index variable. Additionally, as a part of the 
lowering process, \Treebeard{} determines the working set of the loop with caching enabled 
and generates a caching operation with the appropriate limits.

Each of the caching operations take parameters that specify the set of trees or rows that need 
to be cached. The caching operations are defined as follows. 
\begin{itemize}
  \item \textbf{\op{cacheTrees(forest, start, end)}:} This operation caches the trees in ensemble
  \op{forest} from \op{start} to \op{end}. The trees are cached in the order in which they are
  specified in the ensemble. 
  \item \textbf{\op{cacheRows(data, start, end)}:} This operation caches the rows in the input 
  array \op{data} from \op{start} to \op{end}. The rows are cached in the same order as in the
  input array.
\end{itemize}

\subsection{Lowering of Caching Operations}
When the MIR is lowered to LIR, the cache ops are lowered to target-specific code. Each of 
the two caching operations is lowered differently for the CPU and the GPU.

For the \op{cacheRows} operation, \Treebeard{} uses pre-implemented lowerings for both
CPU and GPU. This is possible because the input is currently assumed to be a dense
array format. Therefore, regardless of any other configuration choices (like what representation 
is used for the model itself), the lowering of the \op{cacheRows} operation is the same. 
For the CPU, the \op{cacheRows} operation is lowered to prefetches. For the GPU, a 
series of coalesced loads read the rows into shared memory.

For the \op{cacheTrees} operation, the lowering is representation-specific. Each representation
provides a lowering to the target-specific code generator to lower the \op{cacheTrees} op when
that representation is used. However, the \Treebeard{} infrastructure does provide helpers 
to generate caching code to cache contiguous regions of memory. These helpers are reused 
as required across different representations.

>>>>

\section{Model Representations}
\label{sec:representations}

The design of the \Treebeard{} compiler allows the implementation of different strategies for 
the in-memory representation of the model. The compiler currently has implementations for 
the three representations shown in Figure \ref{Fig:Representations}. The array and sparse 
representations are the ones described in the \TreebeardOLD{} paper\cite{Treebeard}. The reorg 
representation is the representation used by the RAPIDs library\cite{FIL}. 
The \textbf{array representation} is the simplest representation where the trees are stored in an array
in level order. The \textbf{sparse representation} stores the trees in a sparse format where 
memory is allocated only for nodes present in the tree and nodes contain pointers to their children. The \textbf{reorg representation}
interleaves the array representation of each tree in the model: all root nodes are stored first, then 
the left children of all the roots and so on. This representation was 
designed to improve memory coalescing when tree nodes are being loaded. 
\begin{figure}[htb]
  \centering
  \includegraphics[width=\linewidth]{figures/representations.png}
  \caption{The three representations supported by \Treebeard{}.}
  \label{Fig:Representations}
\end{figure}

One of the major changes we make to the original design of \TreebeardOLD{} \cite{Treebeard} is to
separate the implementation of representations from the rest of the compiler. This allows
us to implement representations as plugins to the compiler. We define an interface 
that representations implement. The code generator is implemented using
this interface thus hiding details of the actual representation from the core 
compiler. Curcially, the interface abstracts how and what buffers 
are allocated, how to move from a node to its child, how trees 
are cached, reading the value of leaves and now threshold and 
feature indices are read from the allocated buffers.
% The interface abstracts the following details. 
% \begin{itemize}
%   \item \textbf{Buffer allocation:} The representation object exposes methods that generate 
%   buffer allocations in the IR. 
%   \item \textbf{Moving to child nodes:} Given the value of the predicate at a node (or tile),
%   generate code to move to the appropriate child node of the current node.
%   \item \textbf{Leaf representation:} The interface abstracts determining whether the current
%   node is a leaf (which is needed to terminate walks) and how to get the value of the leaf
%   (which is needed to get a tree's prediction).
%   \item \textbf{Caching trees:} Since reading the trees into shared memory on GPU (or prefetching)
%   them on CPU require knowledge of the buffer layout, the task of generating caching code for trees
%   is delegated to the representation object. 
%   \item \textbf{Loading thresholds and feature indices:} The representation object provides MLIR 
%   rewrite patterns to lower operations that load thresholds and feature indices to LLVM IR. This
%   is necessary because the details of how to load the thresholds and feature indices from the model 
%   buffers is representation-specific.
% \end{itemize}

In summary, the representation interface abstracts the details of how the model is stored in memory
and allows the compiler to generate code without having to explicitly know the details of the
representation. This design allows us to implement new representations without changing the core
compiler infrastructure. Implementing the representations as plugins also allows us to reuse
the implementations across different lowering pipelines.

% For example, the code for the array 
% and sparse representations is almost fully shared between the CPU and GPU lowering pipelines.

\section{Caching}
\label{sec:caching}

@@

\Treebeard{} provides mechanisms to cache both trees and input rows 
on both the CPU and GPU. As described in Section \ref{sec:schedule}, 
the user can specify that the working set of an iteration of a loop
needs to be cached using the \op{cache} directive. This 
provides a unified way to specify caching of both trees and input rows.
\Treebeard{} implements caching at the granularity of a tree or a row.

% \subsection{IR Representation of Caching}
Caching is encoded in the mid-level IR using the \op{cacheTrees} and \op{cacheRows} operations (Table \ref{Tab:IRSpecification}). 
These operations are generated when the HIR lowered to MIR and \op{cache} is specified on an 
index variable in the schedule. While the HIR is being lowered and a cached index variable is 
encountered, the compiler generates a \op{cacheTrees} or \op{cacheRows} operation depending on 
whether the index variable is a tree or a batch index variable.
\Treebeard{} also determines the working set of the loop  
and generates a caching operation with the appropriate limits.

% Each of the caching operations take parameters that specify the set of trees or rows that need 
% to be cached. The caching operations are defined as follows. 
% \begin{itemize}
%   \item \textbf{\op{cacheTrees(forest, start, end)}:} This operation caches the trees in ensemble
%   \op{forest} from \op{start} to \op{end}. The trees are cached in the order in which they are
%   specified in the ensemble. 
%   \item \textbf{\op{cacheRows(data, start, end)}:} This operation caches the rows in the input 
%   array \op{data} from \op{start} to \op{end}. The rows are cached in the same order as in the
%   input array.
% \end{itemize}

%\subsection{Lowering of Caching Operations}
When the MIR is lowered to LIR, the cache ops are lowered to target-specific code. Each of 
the two caching operations is lowered differently for the CPU and the GPU. On CPU, the cahce operations are lowered to preteches while
on the GPU they are lowered to reads into shared memory.

Lowering the \op{cacheRows} operation is straightforward
% \Treebeard{} uses pre-implemented lowerings for both
% CPU and GPU. This is possible 
because the input is currently assumed to be a dense
array format. The lowering for the \op{cacheRows} operation is implemented directly in the \Treebeard{} compiler.

For the \op{cacheTrees} operation, the lowering is representation-specific. Each representation
provides a lowering to the target-specific code 
generator to lower the \op{cacheTrees} op when
that representation is used.
%  However, the \Treebeard{} infrastructure does provide helpers 
% to generate caching code to cache contiguous regions of memory. These helpers are reused 
% as required across different representations.
